Unifying intra- and inter-specific variation in tropical tree mortality

### 2018-08-08 ms/supplementary.Rnw
Some of the changes suggested by @dfalster
Moderate restructue

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle}

\title{
{\normalfont \fontsize{12pt}{16pt}
{\color{black50} Supporting information for:}\\} \vskip6pt
\fontsize{16pt}{20pt} Unifying intra- and inter-specific variation in tropical tree mortality
}

\author[a,b,1]{James S Camac}
\author[c]{Richard Condit}
\author[d]{Richard G FitzJohn}
\author[e]{Lachlan McCalman}
\author[e]{Daniel Steinberg}
\author[a]{Mark Westoby}
\author[f]{S Joseph Wright}
\author[a,g]{Daniel S Falster}

\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, New South Wales, Australia}
\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Melbourne, Victoria, Australia}
\affil[c]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
\affil[d]{Imperial College, London, United Kingdom}
\affil[e]{Data61, CSIRO, Australia}
\affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney, New South Wales, Australia}

% Please give the surname of the lead author for the running footer
\leadauthor{Camac}

\dates{This supplement was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\hypersetup{
     linkcolor=pnasbluetext
}

\RequirePackage[colorlinks=true, allcolors=blue]{hyperref}

@@

@@

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thesection}{S\arabic{section}}

% \setcounter{secnumdepth}{0}

@@

\section{Estimating True Growth}

@@

\section{Parameter distributions}

@@

\section{Numerical optimisation}

@@

\section{Example Stan code}

@@

\section{References}
\bibliography{refs.bib}

@@

@@

\end{document}

>>>>

\documentclass[9pt,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnassupportinginfo}

% Use the lineno option to display guide line numbers if required.

\title{Unifying intra- and inter-specific variation in tropical tree mortality}
\author{J.S. Camac, R. Condit, R.G. FitzJohn, L. McCalman, D. Steinberg, M. Westoby, S.J. Wright, D.S. Falster}
\correspondingauthor{Corresponding James S Camac.\\E-mail: james.camac@gmail.com}

\usepackage{longtable}

@@

%% Comment/remove this line before generating final copy for submission
%%\instructionspage

@@

%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
\SItext

@@

\section*{An overview of methods for estimating mortality}

While a variety of statistical methods have been applied to model tree mortality, these vary in two key ways. First, some methods model binary survival outcomes directly, while others seek to estimate the underlying (time-dependent) mortality rate experienced by a plant. Estimating hazard rates has the advantage that the model naturally incorporates census interval, and so can be used to make predictions across different time intervals.

Second, models differ in the way the hazard or survival curve responds to covariates. Some studies estimate average mortality rates for populations of each species, thereby quantifying differences between species but ignoring variation among individuals within each species. Most other studies use a model structure which in one way or another allows mortality or survival to decline to zero.

Other studies, model hazard rates using either a proportional hazards structure, such as a Cox proportional-hazard model or complementary-log-log link binomial regression \cite{Kleinbaum:2012tn}, or otherwise directly as an negative exponential function of growth rate. In all these cases, large growth rates cause predicted mortality to decline to zero, such that fast growing individuals never die (Fig 1b). The third approach models survival outcomes rather mortality rates per se, using a logistic function (either via a logistic regression, or GLM with logit link).

Two difficulties arise when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make accurate inferences into how various factors affect tree mortality rates \cite{Poorter:2008iu}. This requirement arises in part because the observed outcome -- alive vs. dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the mortality (or `hazard`) rate for individual $i$ at time $t$. In classic survival analysis \cite{Kleinbaum:2012tn, Zens:2003jv}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed:
\begin{equation} \label{eq:mortality_model}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ is the cumulative hazard an individual experiences from  $t_1$ and $t_2$\cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then a realisation of this cumulative hazard on a probability scale:
\begin{equation}
S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
\end{equation}
Second, as trees are long-lived and we are estimating the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required. Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.

outcomes are modelled using a logistic function (either via a logistic regression, or

An individual's total hazard at a given time point is therefore the sum of these two independent hazards. However, to date, most statistical models of tree mortality do not explicitly estimate the independent effects of these two fundamental hazards faced by plants. Instead, plant mortality is typically modelled using: 1) logistic regression (i.e. with logit link), which does not estimate a rate, but rather the log odds of death at a given time point; or 2) via a proportional hazards approach such as a Cox proportional-hazard model or complementary-log-log link binomial regression, both of which assume hazards are multiplicative, and therefore, not independent of one another \cite{Kleinbaum:2012tn}.

While some methods (like logistic regression) model survival outcomes directly, it preferable to estimate a hazard rate, as these can be rescaled across different time intervals \cite{Kleinbaum:2012tn, Zens:2003jv}.

Second, as trees are long-lived and we are estimating the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required. Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.

Previous approaches for estimating $\lambda_i(t)$ have applied either an average rate per species, or a multiplicative model implying an decline to zero as growth rate 
Various

Second, as trees are long-lived and we are estimating the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required. Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.

\section*{Estimating True Growth}

@@

\section*{Estimating species light demand}

A species' light demand was calculated using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995. The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m). The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix (see also Fig. \ref{figS2}).

For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap). As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots. Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1. This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots. These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots.

\section*{Parameter distributions}

@@

\section*{Numerical optimisation}

@@

\section*{Example Stan code}

@@

\bibliography{refs.bib}

@@

<<spp-table,results="asis",center=TRUE, echo= FALSE,>>= 
tableS2(spp_params)
@

@@

\clearpage
\begin{figure*}[!hb]
\centering
\includegraphics{figures/supp_fig6}
\caption{Pairwise comparison between species' wood density, light demand and maximum dbh.}
\label{figS6}
\end{figure*}
\end{document}
